---
title: "TO404HW1_Final"
author: "Danielle Bidigare, in partnership with Jeffrey Sondheimer"
date: "11/1/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
nyc<- read.csv("nycflights.csv")
```

```{r}
nyc$carrier <- as.factor(nyc$carrier)
nyc$flight <- as.factor(nyc$flight)
nyc$tailnum <- as.factor(nyc$tailnum)
nyc$origin <- as.factor(nyc$origin)
nyc$dest <- as.factor(nyc$dest)
```

## Data Exploration

### How many airlines are there?

```{r}
total_num_airlines <- length(levels(nyc$carrier))
total_num_airlines
```

**There are `r total_num_airlines` airlines**

### How many flights there were by the airline with code `OO`?

```{r}
airline_code_OO <- nrow(nyc[nyc$airline == "OO",])
airline_code_OO
```
**There are `r airline_code_OO` flights with the code OO.**

### How long is the shortest flight out of any NYC airport?

```{r}
shortest_flight <- min(nyc$air_time, na.rm=TRUE)
shortest_flight
```
**The shortest flight out of any NYC airport is `r shortest_flight` minutes.**

### How many flights where there by United Airlines (code: UA) on Jan 12th 2013?

```{r}
jan_12_2013_UAFlights <- nrow(nyc[nyc$carrier=="UA" & nyc$year==2013 & nyc$month==1 & nyc$day==12,])
jan_12_2013_UAFlights
```
**On January 12th, 2013, there were `r jan_12_2013_UAFlights` United Airlines flights.**

## Arrival Delay

### What was the average arrival delay for all airports and all airlines combined in Jan 2013?

```{r}
average_arrival_delay_Jan2013 <- round(mean(nyc$arr_delay[nyc$year==2013 & nyc$month==1], na.rm=TRUE),2)
average_arrival_delay_Jan2013
```
**The average arrival delay for all airports and all airlines combined in January 2013 is `r average_arrival_delay_Jan2013` minutes.**

### What was the median arrival delay for all airports and all airlines combined in Jan 2013?

```{r}
median_arrival_delay_Jan2013 <- median(nyc$arr_delay[nyc$year==2013 & nyc$month==1],na.rm=TRUE)
median_arrival_delay_Jan2013
```
**The median arrival delay for all airports and all airlines combined in January 2013 was `r median_arrival_delay_Jan2013` minutes.**

## Airline Performance 

### Average arrival delays by airline

```{r}
average_arr_delay_airline <- round(tapply(nyc$arr_delay, nyc$carrier, mean, na.rm=TRUE), 2)
average_arr_delay_airline
sort(average_arr_delay_airline, decreasing = FALSE)
```

Barplot of these average arrival delays by airline:
```{r}
barplot(sort(average_arr_delay_airline, decreasing = FALSE), las=2)
```

### Which airline has the highest average arrival delay? Which airline has the smallest average arrival delay? Are there airlines that actually have negative average delay?

```{r}
highest_avg_airline_delay <- names(average_arr_delay_airline[average_arr_delay_airline==max(average_arr_delay_airline)])
highest_avg_airline_delay

smallest_avg_airline_delay <- names(average_arr_delay_airline[average_arr_delay_airline==min(average_arr_delay_airline)])
smallest_avg_airline_delay

negative_avg_airline_delay <- names(average_arr_delay_airline[average_arr_delay_airline<0])
negative_avg_airline_delay
```

**`r highest_avg_airline_delay` has the highest average arrival delay, and `r smallest_avg_airline_delay` has the smallest average arrival delay. `r negative_avg_airline_delay` have negative average airlines delays.**

## Air Gain

The following is a histogram of airgain:
```{r}
nyc$airgain <- nyc$dep_delay-nyc$arr_delay
summary(nyc$airgain)
mean(nyc$airgain, na.rm=TRUE)
median(nyc$airgain, na.rm=TRUE)
str(nyc$airgain)
hist(nyc$airgain)
```

### Do airlines actually gain any time when in air on average?

```{r}
average_airgain <- round(mean(nyc$airgain, na.rm=TRUE),2)
average_airgain
```
**On average, airlines gain `r average_airgain` minutes in air.** 

### Average airgain for different airlines:

```{r}
avg_airgain_by_airline <- round(tapply(nyc$airgain,nyc$carrier,mean,na.rm=TRUE),2)
sort(avg_airgain_by_airline)
best_airgain <- head(sort(names(avg_airgain_by_airline)))
worst_airgain <- tail(sort(names(avg_airgain_by_airline)))
```
**`r best_airgain` are the airlines that did the best job gaining airtime, while `r worst_airgain` are the airlines that did the worst job gaining airtime.**

Below is a barplot to depict the best and worst airlines for airgain:
```{r}
barplot(sort(avg_airgain_by_airline, decreasing = FALSE),las=2)
```

